Community Analysis

### 2022-07-06 Manuscript/manuscript.Rmd
start to update text and figures with new methods, results

```{r initialize-packages}
library(tidyverse)
library(broom)
library(broman)
library(morris)
library(phyloseq)
library(ggpubr)
library(knitr)
library(corncob)
library(ggthemes)
library(microViz)
library(captioner)
library(knitr)
library(kableExtra)
library(flextable)
figures <- captioner()
tables <- captioner(prefix = "Table")
```

@@

Statistical analyses were performed in 'R' [@rcoreteam2018]. To test
whether there was a significant change in CH~4~ oxidation rate ($k$) as a
response to selection, we tested a difference in slopes between the positive
and neutral selection lines. The CH~4~ oxidation rates were strongly 
right-skewed with most values close to zero and few large, positive values. This
resulted in residuals that did not meet the assumptions of constant variance
and normal distribution. Therefore, CH~4~ oxidation rates were log~10~
transformed to better meet the assumptions of a linear model and to make
figures easier to interpret. First, we tested the effect of treatment by
fitting two nested models with and without treatment using 'lm'. We compared these
models using the likelihood ratio test with the 'anova' function. We
then fit the full model to estimate the slope of the positive line, which
represents the change in CH~4~ oxidation rate per passage as a response to
selection.

To estimate the proportion of variance in ecosystem function due to variation in
the microbiome, we estimated narrow-sense heritability ($h^2$) as the
regression of mid-offspring on mid-parent [@falconer1996]. The mid-parent was the
mean for all three selected jars and the mid-offspring was the mean for all jars
produced by those parents. First, we tested if there was an effect of treatment
on the heritability estimate. We compared these models using the likelihood ratio
test with the 'anova' function. We then fit the full model to estimate the
heritability of the Positive and Neutral treatments.

Richness estimates and tests were performed using 'breakaway' [@willis2015]. We
tested a difference in richness between Passages 2 and 5 with both treatments
combined. Then we subset the samples from Passage 5 and compared richness between
the Positive and Neutral treatment. Next, we compared beta-diversity as both
Aitchison dissimilarity, which is the Euclidean distance after center-log ratio
transformation, and the binary Jaccard distance for presence or absence [@aitchison1982]. 
We tested a difference in centroid and dispersion of beta
diversity between the Positive and Neutral treatment and Passage 2 and 5 with
permutational analysis of variance (PERMANOVA) using the 'adonis2' function from
'vegan' [@mcardle2001; @oksanen2019].

To identify taxa that responded to selection on CH~4~ oxidation rate, we tested
differential abundance using a beta-binomial model [@martin2020]. We first
grouped ASVs at the family level keeping ASVs that lacked a family-level taxonomic
assignment. We
then subset the samples in Passage 5 and removed all families with a prevalence
of less than 10\%. We compared families in passage 5 between the Positive and
Neutral treatment to identify taxa that were enriched or depleted as a response
to selection. we used the likelihood ratio test to estimate p-values with a
significance threshold of false-discovery rate < 0.05 [@martin2020]. Estimates
are presented as the expected relative abundance in the positive treatment
relative to the neutral treatment.

@@

We observed a response to selection on whole-ecosystem soil CH~4~ oxidation
rate (`r figures("response", display = "cite")`; `r print_lrt(compare_response_mods, 2, 'LRT of nested models with and without treatment')`. At the start of the experiment, the Positive
treatment had a mean CH~4~ oxidation rate that was
`r percent_less(initial = tidy(response_model)[[1, 2]], final = tidy(response_model)[[3, 2]] + tidy(response_model)[[1, 2]])`\% lower than the
Neutral treatment `r print_lm(response_model, 3, 'difference of y-intercepts')`. 
There was no change in CH~4~ oxidation rate of the Neutral
treatment over the five passages `r print_lm(response_model, 2)`. By contrast,
the Positive treatment had a `r round(per_slope)`\% increase in CH~4~
oxidation rate per passage `r print_lm(response_model, 4)`.

@@

Richness of ASVs decreased between Passage 2 (median = `r myround(rare_median[1, 2], 1)`, IQR = `r myround(rare_median[1, 3], 1)`) and 5 (median = `r myround(rare_median[2, 2], 1)`, IQR = `r myround(rare_median[2, 3], 1)`; Kruskal-Wallace test: chi-squared = 
`r myround(kt[1,1], 1)`, df = `r kt[1,3]`, p < 0.001; Pairwise Wilcoxon test: p < 0.001). However, there was no
difference in richness between the Positive and the Neutral treatment in
Passage 2 or 5 (Pairwise Wilcoxon test: p = `r myround(wt[1, 3], 1)`).  In addition, there was no
correlation between richness and CH~4~ oxidation rate across all passages
and treatments (Spearman's rho = `r myround(s[1,1], 1)`, p = `r myround(s[1,3], 1)`).

@@

`r figures("beta", "Principal component analysis plot of beta diversity for all jars. Dissimilarities are based on Aitchison dissimilarity, which is Euclidean distance after center-log ratio transformation. Colors represent treatment and shapes represent passage number. The ratio of the axes is proportional to the variance explained by each principal component in order to accurately represent the distance between the points.")`

@@

`r figures("da", "Differentially abundant family-level taxa based on a beta-binomial model. Values on the x-axis are estimates and standard errors of the expected difference in the logit-transformed relative abundance between the two treatments. Positive values are enriched in the Positive treatment and negative values are enriched in the Neutral treatment. Taxa presented here are significant at FDR < 0.5")`

>>>>

```{r initialize-packages}
library(tidyverse)
library(broom)
library(broman)
library(morris)
library(phyloseq)
library(ggpubr)
library(knitr)
library(corncob)
library(ggthemes)
library(microViz)
library(captioner)
library(knitr)
library(kableExtra)
library(flextable)
library(ggtext)
figures <- captioner()
tables <- captioner(prefix = "Table")
```

@@

Statistical analyses were performed in 'R' [@rcoreteam2018]. To test
whether there was a significant change in CH~4~ oxidation rate ($k$) as a
response to selection, we tested a difference in slopes between the positive
and neutral selection lines. The CH~4~ oxidation rates were strongly 
right-skewed with most values close to zero and few large, positive values. This
resulted in residuals that did not meet the assumptions of constant variance
and normal distribution. Therefore, CH~4~ oxidation rates were log~10~
transformed to better meet the assumptions of a linear model and to make
figures easier to interpret. First, we tested the difference of slopes between
the positive and neutral treatments by
fitting two nested models with and without the interaction between treatment
and passage using 'lm'. We compared these
models using an F-test with the 'anova' function. We
then fit the full model with the interaction term to estimate the slopes for each treatment, which
represented the change in CH~4~ oxidation rate per passage as a response to
selection.

To estimate the proportion of variance in CH~4~ oxidation rate due to variation in
the microbiome, we estimated narrow-sense heritability ($h^2$) as the
regression of mid-offspring on mid-parent [@falconer1996]. The mid-parent was the
mean for all three selected jars and the mid-offspring was the mean for all jars
produced by those parents. First, we tested if there was an effect of treatment
on the heritability estimate. We compared these models using an F-test
test with the 'anova' function. We then fit the full model to estimate the
heritability of the Positive and Neutral treatments.

Richness was estimated using the method from @hurlbert1971 with a subsample size of `r min_seqs` using the 'rarefy'
function in 'vegan' [@oksanen2019]. We
tested a difference in richness between Passages 2 and 5 with both treatments
combined. Then we subset the samples from Passage 5 and compared richness between
the Positive and Neutral treatment. Next, we compared beta-diversity as the 
Bray-Curtis dissimilarity by averaging random subsets with a subsample size of `r min_seqs` using the 'avgdist' function in 'vegan' [@bray1957; @oksanen2019].
We tested a difference in centroid and dispersion of beta
diversity between the Positive and Neutral treatment and Passage 2 and 5 with
permutational analysis of variance (PERMANOVA) using the 'adonis2' function from
'vegan' [@mcardle2001; @oksanen2019].

To identify taxa that responded to selection on CH~4~ oxidation rate, we tested
differential abundance between the two treatments in passage 5. We first
grouped ASVs at the family level keeping ASVs that lacked a family-level taxonomic
assignment by grouping them at a higher level. We
then subset the samples in Passage 5 and removed all families with a prevalence
of less than 10\% in either treatment. We used three methods for testing differential abundance: ANCOM-II, ALDEx2, and CORNCOB. We then
pulled out the consensus taxa that were significant
with all three tests and plotted their relative abundances. For ANCOM-II, we used the 'ancom' function in the 'ANCOM-BC' package with a cutoff of W = 0.7 [@kaul2017; @lin2020]. For ALDEx2, we used the 'aldex' function in the 'ALDEx2' package with the Welch's t test and we used an effect size of
1 as our significance threshold [@fernandes2014]. Finally, we used CORNCOB with the 'differentialTest' function in the 'corncob' package with the Wald test and without bootstrapping [@martin2020].

@@

We observed a response to selection on whole-ecosystem soil CH~4~ oxidation
rate (`r figures("response", display = "cite")`; difference of slopes:
F~`r compare_response_mods[2,3]`,`r compare_response_mods[2,1]`~ = `r myround(compare_response_mods[2,5], 2)`, p = `r myround(compare_response_mods[2,6], 2)`). At the start of the experiment, the Positive
treatment had a mean CH~4~ oxidation rate that was
`r percent_less(initial = tidy(response_model)[[1, 2]], final = tidy(response_model)[[3, 2]] + tidy(response_model)[[1, 2]])`\% lower than the
Neutral treatment `r print_lm(response_model, 3, 'difference of y-intercepts')`. 
There was no change in CH~4~ oxidation rate of the Neutral
treatment over the five passages `r print_lm(response_model, 2)`. By contrast,
the Positive treatment had a `r round(per_slope)`\% increase in CH~4~
oxidation rate per passage `r print_lm(response_model, 4)`.

@@

Richness of ASVs decreased between Passage 2 (median = `r myround(rare_median[1, 2], 1)`, IQR = `r myround(rare_median[1, 3], 1)`) and 5 (median = `r myround(rare_median[2, 2], 1)`, IQR = `r myround(rare_median[2, 3], 1)`; Kruskal-Wallace test: $\chi^2$ = 
`r myround(kt[1,1], 1)`, df = `r kt[1,3]`, p < 0.001; pairwise Wilcoxon test: p < 0.001). However, there was no
difference in richness between the Positive and the Neutral treatment in
Passage 2 or 5 (pairwise Wilcoxon test: p = `r myround(wt[1, 3], 1)`).  In addition, there was no
correlation between richness and CH~4~ oxidation rate across all passages
and treatments (Spearman's rho = `r myround(s[1,1], 1)`, p = `r myround(s[1,3], 1)`).

@@

`r figures("beta", "Principal component analysis plot of beta diversity for all jars. Dissimilarities are based on rarefied Bray-Curtis dissimilarity. Colors represent treatment and shapes represent passage number.")`

@@

`r figures("da", "Differentially abundant family-level taxa identified by ANCOM-II, ALDEx2, and CORNCOB. Values on the x-axis are relative abundances on a log~10~ scale. Points are colored by treatment. Large points with error bars are the median relative abundance and interquartile range. Taxa in the top panel are enriched in the Positive treatment and taxa in the bottom panel are depleted in the Positive treatment. Taxa are sorted by their effect size with taxa at the top having the largest positive effect size and taxa at the bottom with the largest negative effect size.")`

### 2022-07-06 R/02-process-herit.Rmd
start to update text and figures with new methods, results

---
title: "Process Heritability Data"
author: "Andrew Morris"
date: "`r Sys.Date()`"
output: html_document
---

This script takes in the flux data from script 01 and the data on which jars
were selected in each passage. It then aligns these data so that for each
passage there is a mean flux for all parents, the selected parents, and all
offspring.

>>>>

---
title: "Process Heritability Data"
author: "Andrew Morris"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = TRUE)
knitr::opts_knit$set(root.dir=normalizePath('../'))

```
This script takes in the flux data from script 01 and the data on which jars
were selected in each passage. It then aligns these data so that for each
passage there is a mean flux for all parents, the selected parents, and all
offspring.

### 2022-07-06 R/03-response.Rmd
start to update text and figures with new methods, results

```{r compare_models}
# Compare interaction to no interaction
compare_response_mods <- anova(no_treat_model, response_model, test = 'LRT')
kable(tidy(compare_response_mods))
```

>>>>

```{r compare_models}
# Compare interaction to no interaction
compare_response_mods <- anova(no_treat_model, response_model)
compare_response_mods <- tidy(compare_response_mods)
kable(compare_response_mods)
```

### 2022-07-06 R/08-community-analysis.Rmd
start to update text and figures with new methods, results

@@

@@

s5 <-
  rare_data %>% 
  filter(passage == "G5") %>% 
  cor.test(~ ch4 + rare, data = ., method = "spearman", alternative = "two.sided")
save(kt, wt, rare_median, s, file = "Output/richness_models.RData")

@@

all_taxa$Family <- factor(all_taxa$Family, levels = sort_order$Family)
all_iqm$Family <- factor(all_iqm$Family, levels = sort_order$Family)

p <-
	all_taxa %>%
	ggplot(aes(x = rel_abund, y = Family, color = treat)) +
	geom_point(position=position_jitterdodge()) +
	geom_pointrange(aes(x = median, xmax = third_quartile, xmin = first_quartile, fill = treat), 
			data = all_iqm, 
			position = position_dodge(width = 0.75), 
			shape = 21, 
			color = "black") +
	theme_bw() +
	theme(axis.text.y = element_markdown()) +
	labs(x = "Relative Abundance (%)", y = NULL) +
	scale_color_manual(name = "Treatment",
			   labels = c('Neutral', 'Positive'),
			   values = c('gray40', 'darkorange2')) +
	scale_fill_manual(name = "Treatment",
			   labels = c('Neutral', 'Positive'),
			   values = c('gray40', 'darkorange2'))

p_log10 <- 
	p +
	scale_x_log10(breaks = c(0.001, 0.01, 0.1, 1),
		      labels = c(0.001, 0.01, 0.1, 1))

p
p_log10

@@

ggsave('Manuscript/fig4.jpg', p_log10, width = 7, height = 7)

## All
# ggsave("shared_taxa.png", p)
ggsave("shared_taxa_log.png", p_log10, width = 7, height = 7)

>>>>

min_seqs <- seq_range[1]

@@

```{r compare_richness_methods}

aes_df_long <-
  aes_df %>% 
  as_tibble(rownames = "sample") %>% 
  pivot_longer(starts_with("ASV")) %>%     
  group_by(sample) %>% 
  uncount(value)

subsample <- function(data, sample_size) {
  data %>% 
    sample_n(sample_size) %>% 
    summarize(s = n_distinct(name))
}

empirical <- map_dfr(1:10, ~subsample(aes_df_long, seq_range[1]), .id = "iters")
empirical <-
empirical %>% 
  group_by(sample) %>% 
  summarize(s = mean(s))

inner_join(rare, empirical, by = "sample") %>% 
  ggplot(aes(x = rare, y = s)) +
  geom_abline(intercept = 0, slope = 1, color = "gray") +
  geom_point()
```

This plot shows the correlation between Hurlbert's (1971) approach to
richness estimation (rare) and empirically rarefying 10 times and taking
the mean richness (s).

@@

s5 <-
  rare_data %>% 
  filter(passage == "G5") %>% 
  cor.test(~ ch4 + rare, data = ., method = "spearman", alternative = "two.sided")

min_seqs <- paste(substr(as.character(min_seqs), 1, 3), substr(as.character(min_seqs), 4, 6), sep = ",")
save(kt, wt, rare_median, s, min_seqs, file = "Output/richness_models.RData")

@@

all_taxa$Family <- factor(all_taxa$Family, levels = sort_order$Family)
all_iqm$Family <- factor(all_iqm$Family, levels = sort_order$Family)
all_taxa <-
  all_taxa %>% 
  mutate(direction = if_else(effect > 0, 0, 1))
all_iqm <-
all_taxa %>% 
  select(Family, direction) %>% 
  distinct() %>% 
  full_join(all_iqm, by = "Family")
(p <-
	all_taxa %>%
	ggplot(aes(x = rel_abund, y = Family, color = treat)) +
	geom_point(position=position_jitterdodge()) +
	geom_pointrange(aes(x = median, xmax = third_quartile, xmin = first_quartile, fill = treat), 
			data = all_iqm, 
			position = position_dodge(width = 0.75), 
			shape = 21, 
			color = "black") +
	theme_bw() +
	theme(axis.text.y = element_markdown()) +
	labs(x = "Relative Abundance (%)", y = NULL) +
	scale_color_manual(name = "Treatment",
			   labels = c('Neutral', 'Positive'),
			   values = c('gray40', 'darkorange2')) +
	scale_fill_manual(name = "Treatment",
			   labels = c('Neutral', 'Positive'),
			   values = c('gray40', 'darkorange2')) +
    facet_grid(direction ~ ., scale = "free_y", space = "free")  + 
theme(
  strip.background = element_blank(),
  strip.text.y = element_blank()
))

(p_log10 <- 
	p +
	scale_x_log10(breaks = c(0.001, 0.01, 0.1, 1),
		      labels = c(0.001, 0.01, 0.1, 1)))

@@

ggsave('Manuscript/fig4.jpg', p_log10, width = 5, height = 5)

## All
# ggsave("shared_taxa.png", p)
ggsave("shared_taxa_log.png", p_log10, width = 5, height = 5)
